\chapter{Applications}
\label{applications}

A JAX-RS application consists of one or more resources (see Chapter \ref{resources}) and zero or more providers
(see Chapter \ref{providers}).

\section{Configuration}
\label{config}

\textcolor{highlight green}{The resources and providers that make up a JAX-RS application are configured via an
application-supplied subclass of \lstinline{Application}}. An implementation MAY provide alternate mechanisms for
locating resource classes and providers (e.g. runtime class scanning) but use of \lstinline{Application} is the only
portable means of configuration.

\section{Verification}
\label{verification}

Specific application requirements are detailed throughout this specification and the JAX-RS Javadocs. Implementations
MAY perform verification steps that go beyond what it is stated in this document.

A JAX-RS implementation MAY report an error condition if it detects that two or more resources could result in an
ambiguity during the execution of the algorithm described Section \ref{request_matching}. For example, if two resource
methods in the same resource class have identical (or even intersecting) values in all the annotations that are relevant
to the algorithm described in that section. The exact set of verification steps as well as the error reporting mechanism
is implementation dependent.

\section{Publication}

Applications are published in different ways depending on whether the application is run in a Java SE environment or
within a container. This section describes the alternate means of publication.

\subsection{Java SE}

In a Java SE environment a configured instance of an endpoint class can be obtained using the createEndpoint method of
RuntimeDelegate. The application supplies an instance of \lstinline{Application} and the type of endpoint required. An
implementation MAY support zero or more endpoint types of any desired type.

How the resulting endpoint class instance is used to publish the application is outside the scope of this specification.

\subsubsection{JAX-WS}

An implementation that supports publication via JAX-WS MUST support createEndpoint with an endpoint type of
javax.xml.ws.Provider. JAX-WS describes how a \lstinline{Provider} based endpoint can be published in an SE environment.

\subsection{Servlet}
\label{servlet}

A JAX-RS application is packaged as a Web application in a \textcolor{highlight green}{\lstinline{.war}} file. The
application classes are packaged in \textcolor{highlight green}{\lstinline{WEB-INF/classes}} or
\textcolor{highlight green}{\lstinline{WEB-INF/lib}} and required libraries are packaged in
\textcolor{highlight green}{\lstinline{WEB-INF/lib}}. See the Servlet specification for full details on packaging of web
applications.
