\chapter{Validation}
\label{validation}

Validation is the process of verifying that some data obeys one or more pre-defined constraints. The Bean Validation
specification defines an API to validate Java Beans. This chapter describes how JAX-RS provides native support for
validating resource classes. See Section \ref{bv_support} for more information on implementation requirements.

\section{Constraint Annotations}
\label{constraint_annotations}

The Server API provides support for extracting request values and mapping them into Java fields, properties and
parameters using annotations such as @HeaderParam, @QueryParam, etc. It also supports mapping of
request entity bodies into Java objects via non-annotated parameters (i.e., parameters without any JAX-RS annotations).
See Chapter \ref{resources} for additional information.

In earlier versions of JAX-RS, any additional validation of these values needed to be performed programmatically. This
version of JAX-RS introduces support for declarative validation based on the Bean Validation specification.

The Bean Validation specification supports the use of constraint annotations as a way of declaratively validating beans,
method parameters and method returned values. For example, consider the following resource class augmented with
constraint annotations:

\begin{lstlisting}[language=Java]
    @Path("/")
    class MyResourceClass {

    @POST
    @Consumes("application/x-www-form-urlencoded")
    public void registerUser(
    @NotNull @FormParam("firstName") String firstName,
    @NotNull @FormParam("lastName") String lastName,
    @Email @FormParam("email") String email) {
    ...
    }
    }
\end{lstlisting}

The annotations @NotNull and @Email impose additional constraints on the form parameters
firstName, lastName and email. The @NotNull constraint is built-in to
the Bean Validation API; the @Email constraint is assumed to be user defined in the example above.
These constraint annotations are not restricted to method parameters, they can be used in any location in which the
JAX-RS binding annotations are allowed with the exception of constructors and property setters. Rather than using method
parameters, the MyResourceClass shown above could have been written as follows:

\begin{lstlisting}[language=Java]
    @Path("/")
    class MyResourceClass {

        @NotNull
        @FormParam("firstName")
        private String firstName;

        @NotNull
        @FormParam("lastName")
        private String lastName;

        private String email;

        @FormParam("email")
        public void setEmail(String email) {
            this.email = email;
        }

        @Email
        public String getEmail() {
            return email;
        }

        ...
    }
\end{lstlisting}

Note that in this version, firstName and lastName are fields initialized via injection and
email is a resource class property. Constraint annotations on properties are specified in their
corresponding getters.

Constraint annotations are also allowed on resource classes. In addition to annotating fields and properties, an
annotation can be defined for the entire class. Let us assume that @NonEmptyNames validates that one of the
two name fields in MyResourceClass is provided. Using such an annotation, the example above can be extended
as follows:

\begin{lstlisting}[language=Java]
    @Path("/")
    @NonEmptyNames
    class MyResourceClass {

        @NotNull
        @FormParam("firstName")
        private String firstName;

        @NotNull
        @FormParam("lastName")
        private String lastName;

        private String email;

        ...
    }
\end{lstlisting}

Constraint annotations on resource classes are useful for defining cross-field and cross-property constraints.

\section{Annotations and Validators}
\label{annotations_and_validators}

Annotation constraints and validators are defined in accordance with the Bean Validation specification. The
@Email annotation shown above is defined using the Bean Validation @Constraint
meta-annotation:

\begin{lstlisting}[language=Java]
    @Retention(RUNTIME)
    @Target( { METHOD, FIELD, PARAMETER })
    @Constraint(validatedBy = EmailValidator.class)
    public @interface Email {

        String message() default "{com.example.validation.constraints.email}";
        Class<?>[] groups() default {};
        Class<? extends Payload>[] payload() default {};
    }
\end{lstlisting}

The @Constraint annotation must include a reference to the validator class that will be used to validate
decorated values. The \lstinline{EmailValidator} class must implement \lstinline{ConstraintValidator<Email, T>} where
\lstinline{T} is the type of values being validated. For example,

\begin{lstlisting}[language=Java]
    public class EmailValidator implements ConstraintValidator<Email, String> {

        public void initialize(Email email) {
            ...
        }

        public boolean isValid(String value, ConstraintValidatorContext context) {
            ...
        }
    }
\end{lstlisting}

Thus, \lstinline{EmailValidator} applies to values annotated with @Email that are of type
\lstinline{String}. Validators for different types can be defined for the same constraint annotation.

Constraint annotations must also define a \lstinline{groups} element to indicate which processing groups they are
associated with. If no groups are specified (as in the example above) the \lstinline{Default} group is assumed. For
simplicity, JAX-RS implementations are NOT REQUIRED to support processing groups other than \lstinline{Default}. In what
follows, we assume that constraint validation is carried out in the \lstinline{Default} processing group.

\section{Entity Validation}
\label{entity_validation}

Request entity bodies can be mapped to resource method parameters. There are two ways in which these entities can be
validated. If the request entity is mapped to a Java bean whose class is decorated with Bean Validation annotations,
then validation can be enabled using \lstinline{@Valid}:

\begin{lstlisting}[language=Java]
    @StandardUser
    class User {

        @NotNull
        private String firstName;
        ...
        }

    @Path("/")
    class MyResourceClass {

        @POST
        @Consumes("application/xml")
        public void registerUser(@Valid User user) {
            ...
        }
    }
\end{lstlisting}

In this case, the validator associated with \lstinline{@StandardUser} (as well as those for non-class level constraints
like \lstinline{@NotNull}) will be called to verify the request entity mapped to \lstinline{user}. Alternatively, a new
annotation can be defined and used directly on the resource method parameter.

\begin{lstlisting}[language=Java]
    @Path("/")
    class MyResourceClass {

        @POST
        @Consumes("application/xml")
        public void registerUser(@PremiumUser User user) {
            ...
        }
    }
\end{lstlisting}

In the example above, \lstinline{@PremiumUser} rather than \lstinline{@StandardUser} will be used to validate the
request entity. These two ways in which validation of entities can be triggered can also be combined by including
\lstinline{@Valid} in the list of constraints. The presence of \lstinline{@Valid} will trigger validation of all the
constraint annotations decorating a Java bean class. This validation will take place in the \lstinline{Default}
processing group unless the \lstinline{@ConvertGroup} annotation is present. See for more information on
\lstinline{@ConvertGroup}.

Response entity bodies returned from resource methods can be validated in a similar manner by annotating the resource
method itself. To exemplify, assuming both \lstinline{@StandardUser} and \lstinline{@PremiumUser} are required to be
checked before returning a user, the \lstinline{getUser} method can be annotated as shown next:

\begin{lstlisting}[language=Java]
    @Path("/")
    class MyResourceClass {

        @GET
        @Valid
        @PremiumUser
        @Path("{id}")
        @Produces("application/xml")
        public User getUser(@PathParam("id") String id) {
            User u = findUser(id);
            return u;
        }
        ...
    }
\end{lstlisting}

Note that \lstinline{@PremiumUser} is explicitly listed and \lstinline{@StandardUser} is triggered by the presence of
the \lstinline{@Valid}\ annotation -- see definition of \lstinline{User} class earlier in this section.

\section{Default Validation Mode}
\label{default_validation_mode}

Validation is enabled by default only for the so called constrained methods. Getter methods as defined by the Java
Beans specification are not constrained methods, so they will not be validated by default. The special annotation
\lstinline{@ValidateOnExecution} can be used to selectively enable and disable validation. For example, you can enable
validation on method \lstinline{getEmail} shown above as follows:

\begin{lstlisting}[language=Java]
    @Path("/")
    class MyResourceClass {

        @Email
        @ValidateOnExecution
        public String getEmail() {
            return email;
        }

        ...
    }
\end{lstlisting}

The default value for the \lstinline{type} attribute of \lstinline{@ValidateOnExecution} is \lstinline{IMPLICIT} which,
in the example above, results in method \lstinline{getEmail} being validated. See \cite{bv11} for more information on
other uses of this annotation.

Note that if validation for getter methods is enabled and a resource method's signature obeys the rules for getters, the
resource method may be (unintentionally) invoked during validation. Conversely, if validation for getter methods is
disabled and the matching resource method's signature obeys the rules for getters, the JAX-RS runtime will still
validate the method (i.e., the validation preference will be ignored) before invocation.

\section{Annotation Inheritance}
\label{annotation_inheritance}

    It is worth noting that these rules are incompatible with those defined in Section \ref{annotationinheritance}.
    Generally speaking, constraint annotations in are cumulative (can be strengthen) across a given type hierarchy while
    JAX-RS annotations are inherited or, overridden and ignored.

    The goal of this specification is to enable validation of JAX-RS resources by leveraging existing Bean Validation
    implementations.

    \section{Validation and Error Reporting}
    \label{validation_and_error_reporting}

    Constraint annotations are allowed in the same locations as the following annotations: \lstinline{@MatrixParam},
    \lstinline{@QueryParam}, \lstinline{@PathParam}, \lstinline{@CookieParam}, \lstinline{@HeaderParam} and
    \lstinline{@Context}, except in class constructors and property setters. Specifically, they are allowed in resource
    method parameters, fields and property getters as well as resource classes, entity parameters and resource methods
    (return values).

    The default resource class instance lifecycle is per-request in JAX-RS. Implementations MAY support other lifecycles;
    the same caveats related to the use of other JAX-RS annotations in resource classes apply to constraint annotations. For
    example, a constraint validation annotating a constructor parameter in a resource class whose lifecycle is singleton
    will only be executed once.

    JAX-RS implementations SHOULD use the following process to validate resource class instances after they have been
    instantiated:

\begin{description}
    \item[Phase 1] Inject field values and initialize bean properties as described in Section \ref{resource_field}.
    \item[Phase 2] Validate annotations on fields, property getters (if enabled) and the resource class. The order in
    which these validations are executed is implementation dependent.
    \item[Phase 3] Validate annotations on parameters passed to the resource method matched.
    \item[Phase 4] If no constraint violations found thus far, invoke resource method and validate returned value.
\end{description}

The exception model defines a base class javax.validation.ValidationException and a few subclasses to report errors that
are specific to constraint definitions, constraint declarations, group definitions and constraint  violations. JAX-RS
implementations MUST provide a default exception mapper (see Section~\ref{exceptionmapper}) for
javax.validation.ValidationException according to the following rules:

\begin{enumerate}
    \item If the exception is of type javax.validation.ValidationException or any of its subclasses
    excluding javax.validation.ConstraintViolationException, then it is mapped to a response with status code 500
    (Internal Server Error).
    \item If the exception is an instance of javax.validation.ConstraintViolationException, then:
    \begin{enumerate}
        \item If the exception was thrown while validating a method return type, then it is mapped to a response with
        status code 500 (Internal Server Error) \footnote{The property path of a ConstraintViolation provides
        information about the location from which an exception originated. See Javadoc for more information.}.
        \item Otherwise, it is mapped to a response with status code 400 (Bad Request).
    \end{enumerate}
\end{enumerate}

In all cases, JAX-RS implementations SHOULD include a response entity describing the source of the error; however, the
exact content and format of this entity is beyond the scope of this specification. As described in
Section~\ref{exceptionmapper}, applications can provide their own exception mappers and, consequently, customize the
default mapper described above.
